Energy Clustering

### 2017-10-07 energy_clustering/draft/energy_clustering.tex

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
\INPUT Gram matrix $G\in\mathbb{R}^{n\times n}$, and number of clusters $k$
\OUTPUT label matrix $Z \in \{ 0, 1\}^{n \times k}$
\STATE set the diagonal of $G$ to zero: $G_{ii} \leftarrow 0$ 
        for $i=1,\dotsc,n$
\STATE $D\leftarrow\diag(d_1,\dotsc,d_n)$ 
        where $d_i \leftarrow \sum_{j=1}^n G_{ij}$
\STATE solve the generalized eigenvalue problem $G v_i = \lambda_i D v_i$,
        where $i=1,\dotsc,n$ with eigenvalues, and corresponding eigenvectors,
        obeying the ordering
        $\lambda_1 \ge \lambda_2 \ge \dotsc \lambda_n$
\STATE $Y \leftarrow (D v_1| \dotsm | D v_k) 
        \in \mathbb{R}^{n\times k}$ 
has the top $k$ eigenvectors (multiplied by $D$) as columns
\STATE normalize the rows $Y$, i.e. 
$\widetilde{Y}_{ij} \leftarrow Y_{ij} \Big/ 
\left(\sum_{j=1}^n |Y_{ij}|^2\right)^{1/2}$
\STATE apply $k$-means on $\{ y_i \}_{i=1}^n$, 
where $y_i = \widetilde{Y}_{i\bullet}$
is the $i$th row of $\widetilde{Y}$
\STATE $Z_{ij} \leftarrow 1$ if $y_i$ is in cluster $j$, 
        and $Z_{ij} \leftarrow 0$ otherwise
\end{algorithmic}
\caption{
\label{spectralalgo}
$\mathcal{E^S}$-clustering is an exact method to solve
the relaxed problem \eqref{eq:relaxed}.
\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, 
                initial label matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the energy costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
        in each cluster%, obtained from $D=Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
            %for $\ell=1,2,\dots,k$ and $\ell \ne j$; 
            using \eqref{eq:changeQ} \label{stepmove}
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and 
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ G_{ii}$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
$\mathcal{E}^H$-clustering: Hartigan's method for energy clustering.
This algorithm~~~~~~~~ \\ finds local solutions to  
the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

From the results of \cite{Telgarsky}, summarized in the end of the previous
section, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase while
keeping the number of points in each cluster fixed.
Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, yielding an optimal 
accuracy of $\approx 0.86$.
For each $D$ we generate $100$ Monte Carlo runs, 
sampling $100$ points for each cluster.
We apply each algorithm to the resulting dataset 
and compute the accuracy \eqref{eq:accuracy}.
$\mathcal{E}^H$- and $\mathcal{E}^L$-clustering
both use the standard
kernel \eqref{eq:standard_metric}, as $\mathcal{E}^S$-clustering\footnote{We
compared our implementation of spectral clustering provided in
Algorithm~\ref{spectralalgo} with the spectral clustering routine provided
by \emph{scikit-learn} library \cite{scikit-learn}, 
which is based on \cite{Malik}, and in the few
cases considered the algorithms had similar performance.}, 
and we initialize the algorithms
with the initialization procedure of $k$-means++ \cite{Vassilvitskii}.
For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
\cite{scikit-learn}, where $k$-means is also initialized with $k$-means++,
while GMM is initialized with the output of $k$-means. Every algorithm
is ran $5$ times, picking the best objective function value.
We plot the average accuracy versus $D$, with error bars indicating
standard error.
The results are shown in Fig.~\ref{fig:gauss}a, where 
$\mathcal{E}^H$-clustering and $\mathcal{E}^S$-clustering have practically
the same performance in this experiment, which is 
much higher than $\mathcal{E}^L$-clustering (kernel
$k$-means), $k$-means, and GMM. The improvement is noticeable 
specially in high dimensions.
Note that in this setting, $k$-means and GMM are consistent estimators,
however, energy clustering degrades much less as dimension increases.

@@

We now allow the diagonal covariance of the second Gaussian distribution to
have different values by choosing
$(\Sigma_2)_{ii} \stackrel{iid}{\sim} \textnormal{Uni}(1,5)$ for
$i=1,\dotsc,10$, and $(\Sigma_2)_{ii} = 1$ for $i=11,\dotsc,D$. For instance,
we consider the fixed setting
\begin{equation}
\label{eq:gauss2}
\begin{split}
x & \in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = I_D, \quad
\Sigma_2 = \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
1.969, 4.035,   4.237,  2.813,  3.637),
\end{split}
\end{equation}
as we vary $D$. Note that there is nothing special about the numbers in
$\widetilde{\Sigma}_{10}$, it is simply a random choice of sampling
$\textnormal{Uni}(1,5)$ ten times, and one could pick any other choices
as well.
The Bayes error is kept fixed yielding
an optimal accuracy $\approx 0.95$.
In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
$k$-means, and GMM, where we do $100$ Monte Carlo runs, sampling
$100$ points for each cluster. The results of $\mathcal{E}^H$-
and $\mathcal{E}^S$-clustering are superior than the other algorithms, 
specially
in high dimensions.

@@

Besides the standard kernel from energy statistics 
\eqref{eq:standard_metric}, 
consider the following two other semimetrics with their respective generating
kernels:
\begin{align}
\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
 \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
\| x \|^{1/2} + \| y \|^{1/2} 
- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
\rho_{e}(x,y) &= 
2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
 \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
\label{eq:rhoe}
\end{align}
The kernel $K_{1/2}(x,y)$ corresponds to the standard energy distance
with $\alpha=1/2$.
We sample data from the following normal distribution in $20$ dimensions:
\begin{equation}
\label{eq:20gauss}
\begin{split}
x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
\Sigma_2 = I_{20}.
\end{split}
\end{equation}
The optimal accuracy based on Bayes
classification error is $\approx 0.90$. 
%Since previous results should be already convincing enough that
%$\mathcal{E}^H$-clustering, based on Hartigan's method, is superior
%to $\mathcal{E}^L$-clustering, which is based on Lloyd's method, in the
%following 
We sample an equal number of points for each cluster which is progressively
increased, $n \in [5, 200]$. 
The results over $100$ Monte Carlo trials are 
shown in Fig.~\ref{fig:consist}a, where we compare
$\mathcal{E}^H$-clustering, with different
kernels, to $k$-means and GMM.
Note that for small $n$ all methods
are superior than GMM, which slowly catches up and tend to optimal Bayes,
as expected since it is a consistent estimator in this case. 
Note also that $\mathcal{E}^H$-clustering
with kernel $\kk_e$ is as accurate as GMM for large number of points, however,
it is superior for small number of points. Still for
the same setting, in Fig.~\ref{fig:consist}b we
we show the difference in accuracy provided by $\mathcal{E}^H$- minus
$\mathcal{E}^L$- and $\mathcal{E}^S$-clustering, when using the
kernel $\kk_e$.
Note that $\mathcal{E}^H$ was
always superior than the other methods.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{normal_kernels.pdf}\\[-1.0em](a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{lognormal_kernels.pdf}\\[-1.0em](c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{normal_kernels_difference.pdf}\\[-1.0em](b)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{lognormal_kernels_difference.pdf}\\[-1.0em]
(d)
\end{minipage}
\caption{
\label{fig:consist}
$\mathcal{E}^H$-clustering with different kernels versus
$k$-means and GMM. In both settings Bayes accuracy
is $\approx 0.9$. We show average accuracy (error bars are standard error)
versus number of points for $100$ Monte Carlo runs.
(a,b) Normal distribution \eqref{eq:20gauss}. 
(c,d) Lognormal distribution~\eqref{eq:20loggauss}.
The plots in (c) and (d) consider the difference in accuracy
between $\mathcal{E}^H$- versus $\mathcal{E}^L$- and
$\mathcal{E}^S$-clustering.
}
\end{figure}

Now consider the same parameters as in \eqref{eq:20gauss} but with
lognormal distributions, 
\begin{equation}
\label{eq:20loggauss}
\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
\end{equation}
The same previous experiment is shown in 
Fig.~\ref{fig:consist}c.
Energy clustering still performs accurately, with any of the utilized
kernels, providing better results than $k$-means and GMM on
this non-normal data. The kernel $\kk_e$ still provides the best results
for small number of points, but its performance is eventually achieved by
$\kk_{1/2}$, indicating that $\alpha \approx 1/2$ in the standard
energy distance should be more appropriate for skewed distributions.
Still for the same setting, in Fig~.\ref{fig:consist}d
we show the difference in accuracy provided by $\mathcal{E}^H$ versus
$\mathcal{E}^L$ and $\mathcal{E}^S$, for the kernel $\kk_e$. Again,
the accuracy provided by $\mathcal{E}^H$ is higher than the other methods,
although not much higher than $\mathcal{E}^S$ in this example.
The two experiments of Fig.~\ref{fig:consist}
illustrate how energy statistics clustering does not make assumptions
about the distribution of the data, performing well in different settings
with the same kernel.

@@

Consider the following choices of semimetric and 
corresponding generating kernel:
\begin{align}
\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
\| x \|^\alpha +
\| y \|^\alpha -
\| x-y \|^\alpha \right) ,
\label{eq:rhoalpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
\label{eq:rhoesigma} \\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
\label{eq:rhogsigma} 
\end{align}
In Fig.~\ref{fig:other} we have examples of
complex two dimensional datasets. The two parallel cigars in (a)
have $200$ points each. The concentric circles
in (b) and (c) have $400$ points for each class.
We apply $\mathcal{E}^H$-clustering  with different choices
of kernels in \eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. We also consider
the best kernel choice for each example for $\mathcal{E}^S$-clustering. 
Moreover, we consider
$k$-means and GMM. We perform $10$ Monte Carlo runs for each example
and show the results in Table~\ref{table:other}, where the respective 
choice of parameters for the kernels are indicated. For (a) we initialize
all algorithms with $k$-means++, and for (b) and (c) we initialize at
random.
We observe that $\mathcal{E}^H$-clustering has superior performance
in every example, and in particular better than $\mathcal{E}^S$-clustering.
For the data in Fig.~\ref{fig:other}a 
the semimetrics $\rho_1$ and $\rho_{1/2}$ are
still able to provide accurate results, however, for the
examples in (b) and (c) the kernel choice is more sensitive.

\begin{table}[h]
\renewcommand*{\arraystretch}{0.75}
\begin{tabular}{@{}r  l g  l g  l g@{}}
\toprule[1pt]
 & & \emph{Fig.~\ref{fig:other}a}
 & & \emph{Fig.~\ref{fig:other}b}
 & & \emph{Fig.~\ref{fig:other}c} \\
\midrule[0.5pt]
\multirow{4}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
& $\rho_{1}$ & $0.705\pm 0.065$
& $\rho_{1}$ & $0.521\pm 0.005$
& $\rho_{1}$ & $0.366\pm 0.005$ \\
& $\rho_{1/2}$ & $0.952\pm 0.048$
& $\rho_{1/2}$ & $0.522\pm 0.004$
& $\rho_{1/2}$ & $0.367\pm 0.005$ \\
& $\widetilde{\rho}_{2}$ & $\bm{0.9987\pm 0.0008}$
& $\widetilde{\rho}_{1}$ & $0.778\pm 0.075$
& $\widetilde{\rho}_{2}$ & $0.374\pm 0.009$ \\
& $\widehat{\rho}_{2}$  & $0.956\pm 0.020$
& $\widehat{\rho}_{1}$  & $\bm{1.0\pm 0.0}$
& $\widehat{\rho}_{2}$ & $\bm{0.572\pm 0.041}$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$\mathcal{E}$-spectral~~~~}
& $\widetilde{\rho}_{2}$ & $0.557\pm 0.014$ 
& $\widehat{\rho}_{1}$ & $0.732\pm 0.002$ 
& $\widehat{\rho}_{2}$ & $0.370\pm 0.005$  \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$k$-means}~~~~ 
& & $0.550\pm 0.011$
& & $0.522\pm 0.004$
& & $0.362\pm 0.004$ \\
\emph{GMM}~~~~
& & $0.903\pm 0.064$
& & $0.595\pm 0.011$
& & $0.520\pm 0.034$ \\
\arrayrulecolor{black}\bottomrule[1pt]
\end{tabular}
\caption{\label{table:other}
Clustering data from Fig.~\ref{fig:other}.
We show average
accuracy over $10$ Monte Carlo runs.
}
\end{table}

>>>>

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
\INPUT Gram matrix $G\in\mathbb{R}^{n\times n}$, and number of clusters $k$
\OUTPUT label matrix $Z \in \{ 0, 1\}^{n \times k}$
\STATE set the diagonal of $G$ to zero: $G_{ii} \leftarrow 0$ 
        for $i=1,\dotsc,n$
\STATE $D\leftarrow\diag(d_1,\dotsc,d_n)$ 
        where $d_i \leftarrow \sum_{j=1}^n G_{ij}$
\STATE solve the generalized eigenvalue problem $G v_i = \lambda_i D v_i$,
        where $i=1,\dotsc,n$ with eigenvalues, and corresponding eigenvectors,
        obeying the ordering
        $\lambda_1 \ge \lambda_2 \ge \dotsc \lambda_n$
\STATE $Y \leftarrow (D v_1| \dotsm | D v_k) 
        \in \mathbb{R}^{n\times k}$ 
has the top $k$ eigenvectors (multiplied by $D$) as columns
\STATE normalize the rows $Y$, i.e. 
$\widetilde{Y}_{ij} \leftarrow Y_{ij} \Big/ 
\left(\sum_{j=1}^n |Y_{ij}|^2\right)^{1/2}$
\STATE apply $k$-means on $\{ y_i \}_{i=1}^n$, 
where $y_i = \widetilde{Y}_{i\bullet}$
is the $i$th row of $\widetilde{Y}$
\STATE $Z_{ij} \leftarrow 1$ if $y_i$ is in cluster $j$, 
        and $Z_{ij} \leftarrow 0$ otherwise
\end{algorithmic}
\caption{
\label{spectralalgo}
$\mathcal{E^S}$-clustering is an exact spectral method to solve
the relaxed problem \eqref{eq:relaxed}.
\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, 
                initial label matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the energy costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
        in each cluster%, obtained from $D=Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
            %for $\ell=1,2,\dots,k$ and $\ell \ne j$; 
            using \eqref{eq:changeQ} \label{stepmove}
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and 
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$ and
    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ G_{ii}$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
$\mathcal{E}^H$-clustering is Hartigan's method for energy clustering.
This algorithm~~~~~~ \\ finds local solutions to  
the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

@@

From the results of \cite{Telgarsky}, summarized in the end of the previous
section, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase while
keeping the number of points in each cluster fixed.
Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, yielding an optimal 
accuracy of $\approx 0.86$.
For each $D$ we generate $100$ Monte Carlo runs, 
sampling $100$ points for each cluster.
We apply each algorithm to the resulting dataset 
and compute the accuracy \eqref{eq:accuracy}. The algorithms
$\mathcal{E}^H$- and $\mathcal{E}^L$-clustering
both use the standard
kernel \eqref{eq:standard_metric}, as well as the spectral
$\mathcal{E}^S$-clustering method\footnote{We
compared our implementation of spectral clustering provided in
Algorithm~\ref{spectralalgo} with the spectral clustering routine provided
by \emph{scikit-learn} library \cite{scikit-learn}, 
which is based on \cite{Malik}. In the few
cases considered the algorithms had similar performance.}, 
and we initialize the algorithms
with the initialization of $k$-means++ \cite{Vassilvitskii}.
For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
\cite{scikit-learn}, where $k$-means is also initialized with $k$-means++,
while GMM is initialized with the output of $k$-means. Every algorithm
is ran $5$ times, returning the answer with best objective function value.
We plot the average accuracy versus $D$, with error bars indicating
standard error.
The results are shown in Fig.~\ref{fig:gauss}a. We can see that
$\mathcal{E}^H$- and $\mathcal{E}^S$-clustering have practically
the same performance, which is 
much higher than $\mathcal{E}^L$-clustering (kernel
$k$-means). Moreover, $\mathcal{E}^H$-clustering outperforms 
$k$-means and GMM, where the improvement is noticeable specially in 
high dimensions.
Note that in this setting $k$-means and GMM are consistent estimators,
however, energy clustering degrades much less as dimension increases.

@@

We now allow the diagonal covariance of the second Gaussian distribution to
have different values by choosing
$(\Sigma_2)_{ii} \stackrel{iid}{\sim} \textnormal{Uni}(1,5)$ for
$i=1,\dotsc,10$, and $(\Sigma_2)_{ii} = 1$ for $i=11,\dotsc,D$. For instance,
we consider the fixed setting
\begin{equation}
\label{eq:gauss2}
\begin{split}
x & \in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = I_D, \quad
\Sigma_2 = \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
1.969, 4.035,   4.237,  2.813,  3.637),
\end{split}
\end{equation}
as we vary $D$. Note that there is nothing special about the numbers in
$\widetilde{\Sigma}_{10}$, it is simply a random choice of sampling
$\textnormal{Uni}(1,5)$ ten times, and one could pick any other choices
as well.
The Bayes error is kept fixed yielding
an optimal accuracy $\approx 0.95$.
In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
$k$-means, and GMM.
Note that GMM performs better in low dimensions, 
since it is a consistent estimator.
However, $\mathcal{E}^H$ and $\mathcal{E}^S$ degrade much less as dimensions
increase, performing better than $k$-means and GMM in high dimensions.

@@

Besides the standard kernel from energy statistics 
\eqref{eq:standard_metric}, 
consider the following two other semimetrics with their respective generating
kernels:
\begin{align}
\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
 \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
\| x \|^{1/2} + \| y \|^{1/2} 
- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
\rho_{e}(x,y) &= 
2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
 \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
\label{eq:rhoe}
\end{align}
The kernel $K_{1/2}(x,y)$ corresponds to the standard energy distance
with $\alpha=1/2$.
We sample data from the following normal distribution in $20$ dimensions:
\begin{equation}
\label{eq:20gauss}
\begin{split}
x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
\Sigma_2 = I_{20}.
\end{split}
\end{equation}
The optimal accuracy based on Bayes
classification error is $\approx 0.90$. 
We sample an equal number of points for each cluster which is progressively
increased, $n \in [5, 200]$. 
The results over $100$ Monte Carlo trials are 
shown in Fig.~\ref{fig:consist}a, where we compare
$\mathcal{E}^H$-clustering, with different
kernels, to $k$-means and GMM.
Note that for small $n$ all methods
are superior than GMM, which slowly catches up and tend to optimal Bayes,
as expected since it is a consistent estimator in this case. 
Note also that $\mathcal{E}^H$-clustering
with kernel $\kk_e$ is as accurate as GMM for large number of points, however,
it is superior for small number of points. Still for
the same setting, in Fig.~\ref{fig:consist}b we
we show the difference in accuracy provided by $\mathcal{E}^H$ minus
$\mathcal{E}^L$ and $\mathcal{E}^S$, when using the
kernel $\kk_e$.
Note that $\mathcal{E}^H$ was
always superior than the other 
methods (there would be points negative values on the $y$-axis otherwise).

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{normal_kernels.pdf}\\[-1.0em](a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{lognormal_kernels.pdf}\\[-1.0em](c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{normal_kernels_difference.pdf}\\[-1.0em](b)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{lognormal_kernels_difference.pdf}\\[-1.0em]
(d)
\end{minipage}
\caption{
\label{fig:consist}
$\mathcal{E}^H$-clustering with different kernels versus
$k$-means and GMM. In both settings Bayes accuracy
is $\approx 0.9$. We show average accuracy (error bars are standard error)
versus number of points for $100$ Monte Carlo trials.
(a,b) Normal distribution \eqref{eq:20gauss}. 
(c,d) Lognormal distribution~\eqref{eq:20loggauss}.
The plots in (c) and (d) consider the difference in accuracy
between $\mathcal{E}^H$ versus $\mathcal{E}^L$ and
$\mathcal{E}^S$-clustering, with semimetric $\rho_e$.
}
\end{figure}

Now consider the same parameters as in \eqref{eq:20gauss} but with
lognormal distributions, 
\begin{equation}
\label{eq:20loggauss}
\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
\end{equation}
The same previous experiment is shown in 
Fig.~\ref{fig:consist}c.
Energy clustering still performs accurately, with any of the utilized
kernels, providing better results than $k$-means and GMM on
this non-normal data. The kernel $\kk_e$ still provides the best results
for small number of points, but its performance is eventually achieved by
$\kk_{1/2}$, indicating that $\alpha \approx 1/2$ in the standard
energy distance should be more appropriate for skewed distributions.
Still for the same setting, in Fig.~\ref{fig:consist}d
we show the difference in accuracy provided by $\mathcal{E}^H$ versus
$\mathcal{E}^L$ and $\mathcal{E}^S$, for the kernel $\kk_e$. Again,
the accuracy provided by $\mathcal{E}^H$ is higher than the other methods,
although not much higher than $\mathcal{E}^S$ in this example.
The two experiments of Fig.~\ref{fig:consist}
illustrate how energy statistics clustering does not make assumptions
about the distribution of the data, performing well in different settings
with the same kernel.

@@

Consider the following choices of semimetric and 
corresponding generating kernel:
\begin{align}
\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
\| x \|^\alpha +
\| y \|^\alpha -
\| x-y \|^\alpha \right) ,
\label{eq:rhoalpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
\label{eq:rhoesigma} \\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
\label{eq:rhogsigma} 
\end{align}
In Fig.~\ref{fig:other}a--c we have 
complex two dimensional datasets. The two parallel cigars in (a)
have $200$ points each. The concentric circles
in (b) and (c) have $400$ points for each class.
We apply $\mathcal{E}^H$-clustering  with different choices
of kernels in \eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. We also consider
the best kernel choice for each example for the spectral
$\mathcal{E}^S$-clustering. 
Moreover, we consider
$k$-means and GMM. We perform $10$ Monte Carlo runs for each example.
The results are in Table~\ref{table:other}.
%, where the respective 
%choice of parameters for the kernels are indicated. 
For (a) we initialize
all algorithms with $k$-means++, and for (b) and (c) we initialize at
random.
$\mathcal{E}^H$ has superior performance
in every example, and in particular better than the
spectral $\mathcal{E}^S$.
For the data in Fig.~\ref{fig:other}a 
the semimetrics $\rho_1$ and $\rho_{1/2}$ are
still able to provide accurate results, however, for the
examples in (b) and (c) the kernel choice is more sensitive.

\begin{table}[t]
\renewcommand*{\arraystretch}{0.75}
\begin{tabular}{@{}r  l g  l g  l g@{}}
\toprule[1pt]
 & & \emph{Fig.~\ref{fig:other}a}
 & & \emph{Fig.~\ref{fig:other}b}
 & & \emph{Fig.~\ref{fig:other}c} \\
\midrule[0.5pt]
\multirow{4}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
& $\rho_{1}$ & $0.705\pm 0.065$
& $\rho_{1}$ & $0.521\pm 0.005$
& $\rho_{1}$ & $0.393\pm 0.020$ \\
& $\rho_{1/2}$ & $0.952\pm 0.048$
& $\rho_{1/2}$ & $0.522\pm 0.004$
& $\rho_{1/2}$ & $0.486\pm 0.040$ \\
& $\widetilde{\rho}_{2}$ & $\bm{0.9987\pm 0.0008}$
& $\widetilde{\rho}_{1}$ & $0.778\pm 0.075$
& $\widetilde{\rho}_{2}$ & $0.666\pm 0.007$ \\
& $\widehat{\rho}_{2}$  & $0.956\pm 0.020$
& $\widehat{\rho}_{1}$  & $\bm{1.0\pm 0.0}$
& $\widehat{\rho}_{2}$ & $\bm{0.676\pm 0.002}$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$\mathcal{E}$-spectral~~~~}
& $\widetilde{\rho}_{2}$ & $0.557\pm 0.014$ 
& $\widehat{\rho}_{1}$ & $0.732\pm 0.002$ 
& $\widehat{\rho}_{2}$ & $0.364\pm 0.004$  \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$k$-means}~~~~ 
& & $0.550\pm 0.011$
& & $0.522\pm 0.004$
& & $0.368\pm 0.005$ \\
\emph{GMM}~~~~
& & $0.903\pm 0.064$
& & $0.595\pm 0.011$
& & $0.465\pm 0.030$ \\
\arrayrulecolor{black}\bottomrule[1pt]
\end{tabular}
\caption{\label{table:other}
Clustering data from Fig.~\ref{fig:other}.
We show average
accuracy over $10$ Monte Carlo runs.
}
\end{table}
